Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-11 paper.tex
Rough outline of per-field data

\subsection{Case study: Genomics England 100,000 genomes}
In this section we demonstrate the utility of vcfzarr on a large human dataset.
Genomics England’s multi-sample VCF dataset (aggV2) is an 
aggregate of 78,195 gVCFs from rare disease and cancer participants 
recruited as part of the 100,000 Genomes Project~\cite{turnbull2018100}. 
aggV2 comprises approximately 722 million annotated single-nucleotide 
variants (SNVs) and small indels split into 1,371 roughly equal chunks and 
totalling 165.3 Tib of data after \texttt{bgzip} compression. 
The dataset is used for a variety of research purposes, ranging from 
genome-wide association studies (GWAS)~\cite{kousathanas2022whole} and 
imputation~\cite{shi2023genomics} to 
simple queries involving single gene 
regions~\cite{leggatt2023genotype}~\cite{lam2023repeat}.
We converted the data to the intermediate columnar format 
which took X hours with a cluster of Y workers using the 
distributed explode commands. When then converted to Zarr using 
SOMETHING which took X hours using [RESOURCES]. In total, this gave 
X TiB of data, spread across Y thousand files, a compression of 
X over the original VCFs.
The detailed breakdown of per-column compression ratios is 
shown in Table~\ref{tab-genomics-england-data}.

\begin{table}
\begin{tabular}{cc}
Some & data on Genomics England\\
\end{tabular}
\caption{Summary of vcf-zarr performance on the Genomics England 100,000
genomes VCF dataset.
\label{tab-genomics-england-data}}
\end{table}

>>>>

\subsection{Case study: Genomics England 100,000 genomes}
In this section we demonstrate the utility of vcfzarr on a large human dataset.
Genomics England’s multi-sample VCF dataset (aggV2) is an 
aggregate of 78,195 gVCFs from rare disease and cancer participants 
recruited as part of the 100,000 Genomes Project~\cite{turnbull2018100}. 
The dataset comprises approximately 722 million annotated single-nucleotide 
variants (SNVs) and small indels split into 1,371 roughly equal chunks and 
totalling 165.3 TiB of VCF data after \texttt{bgzip} compression. 
The dataset is used for a variety of research purposes, ranging from 
genome-wide association studies (GWAS)~\cite{kousathanas2022whole} and 
imputation~\cite{shi2023genomics} to 
simple queries involving single gene 
regions~\cite{leggatt2023genotype,lam2023repeat}.

We converted the data to the intermediate columnar format 
which took X hours with a cluster of Y workers using the 
distributed explode commands. When then converted to Zarr using 
SOMETHING which took X hours using [RESOURCES]. In total, this gave 
X TiB of data, spread across Y thousand files, a compression of 
X over the original VCFs.
The detailed breakdown of per-column compression ratios is 
shown in Table~\ref{tab-genomics-england-data}.

\begin{table}
\caption{Summary of vcf-zarr performance on the Genomics England 100,000
genomes VCF dataset.
\label{tab-genomics-england-data}}
\begin{tabular}{llccc}
\toprule
Field & type & storage & compress & \%total \\
\midrule
call\_AD &  int16 & 179G & 26 & 24\\
call\_GQ &  int16 & 171G & 13 & 22 \\
call\_DP &  int16 & 141G & 20 & 18.8 \\
call\_FT &  string & 58G & 160 & 7.3 \\
call\_PL &  int16 & 51G & 140 & 6.8 \\
call\_GQX &  int16 & 12G & 190 & 1.6 \\
call\_genotype & int8 & 6.13G & 380 & 0.8 \\
call\_genotype\_mask & bool & 3.66G  & 630 & 0.48\\
call\_genotype\_phased & bool & 3.66G  & 630 & 0.48 \\
variant\_quality & float32 & 22.6M & 2.7 & 0.0003 \\
variant\_allele & string & 21.5M & 11 \\
variant\_position & int32 & 12.7M & 4.7 \\
variant\_ABratio & float32 & 9.6M & 6.3 \\
variant\_AN & int32 & 9.6M & 6.3 \\
variant\_filter & bool & 1.96M & 490 \\
\bottomrule
\end{tabular}
\end{table}
